Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-03-30 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Improve section about Bjordal's variant

\section{A natural deduction proof based on Bj{\o}rdal's axiomatization}
\label{sec:bjordal}

In Bj\o rdal's axiomatization godlikeness is taken as primitive, $G_B$.

\begin{definition}
\label{B:D1}
A property is positive iff it is necessarily possessed by every God-like being.
$$
P_B(\varphi) \equiv \nec \all x. [G_B(x) \imp \varphi(x)]
$$
\end{definition}

\begin{definition}
\label{B:D2}
A maximal composite of an individual's positive properties is a positive property possessed by the individual and necessarily implying every positive property possessed by the individual.
$$MCP(\varphi, x) \equiv (\varphi(x) \wedge P_B (\varphi)) \wedge \all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(\varphi(y) \imp \psi(y)))
$$
\end{definition}

\begin{definition}
\label{B:D3}
Necessary existence of an individual is the necessary exemplification of all its maximal composites.
$$
NE_B (x) \equiv \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))
$$
\end{definition}

\begin{axiom}
\label{A:A1'} If a property is positive, its negation is not positive: 
$$
\all \varphi. [P_B (\varphi) \imp \neg P_B (\neg\varphi)]
$$
\end{axiom}

\begin{axiom}
\label{A:A5'} Necessary existence is a positive property. 
$$
P_B (NE_B )
$$
\end{axiom}

The proof of the possible instantiation of $G_B$ is adjusted by proving Theorem 1 of section \ref{sec:newproof} directly for $G_B$ and not for properties in general. Corollary 1 of section \ref{sec:newproof} follows by proving the positivity of godlikeness as in lemma \ref{B:L1} below.

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(G_B) \imp \pos \ex x.G_B(x)]
$$
\end{theorem}
\begin{proof} \hfill

\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.G_B(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.G_B(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: G_B(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.G_B(x) $} \RightLabel{$\imp_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg G_B(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  G_B(x) \imp \neg G_B(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(G_B(x) \imp \neg G_B(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(G_B(x) \imp \neg G_B(x)) $}\RightLabel{$\wedge_I$}
\UIC{$P(\neg G_B)$}
\AXC{Axiom 6***} \dashedLine
\UIC{$\all \varphi. [P(\neg G_B) \imp \neg P(G_B)]$} \RightLabel{$\all_E$}
\UIC{$ P( \neg G_B(x)) \imp \neg P(G_B) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(G_B) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(G_B)$} 
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.G_B(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.G_B(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(\rho) \imp \pos \ex x.G_B(x) $} \RightLabel{$\all_I$}
\UIC{$ \all \varphi. [P(G_B) \imp \pos \ex x.G_B(x)] $}

\begin{lemma}
\label{B:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G_B(z) \imp \nec \ex x. G_B(x)
$$
\end{lemma}

\begin{small}
\begin{prooftree}
\AXC{$ $}
\UIC{$G_B(x) $} 
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. [G_B(x) \imp G_B(x)] $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $}\RightLabel{ $\wedge_I$}
\BIC{$G_B(x) \wedge P_B(G_B) $}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A5'}}\dashedLine
\UIC{$P_B (NE_B)$} \RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp NE_B(x)]$}\RightLabel{ D\ref{B:D3}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\nec_E$}
\UIC{$\all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\forall_E$}
\UIC{$ [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}
\AXC{$ $} \RightLabel{ $2$}
\UIC{$G_B(x) $}  \RightLabel{ $\imp_E$}
\BIC{$\all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$} \RightLabel{ $\forall_E$}
\UIC{$(MCP(G_B, x) \imp \nec \ex y.G_B(y))]$}
\AXC{$ $}\dashedLine
\UIC{$MCP(G_B, x) $} \RightLabel{ $\imp_E$}
\BIC{$\nec \ex y.G_B(y)$}\RightLabel{ $\imp_I^2$}
\UIC{$G_B(x) \imp \nec \ex y.G_B(y)$}
\end{prooftree}
\end{small}

@@

@@

>>>>

\section{Bj{\o}rdal's Variant in Natural Deduction}
\label{sec:bjordal}

Bj{\o}rdal's variant \citep{bjordal,fuhrmann} is another solution to the Modal Collapse. In striking contrast to other variants, Bj\o rdal's axiomatization takes god-likeness ($G_B$) as a primitive notion, upon which the notion of positive property is defined. This variant also significantly reduces the number of necessary axioms. Scott's variant of Axioms 2, 3 and 4 become unnecessary to prove the final theorem from Bj{\o}rdal's modified definitions 1, 2 and 3 and axioms 1 and 5.

\setcounter{definition}{0}
\begin{definition}[Changed]
\label{B:D1}
A property is \emph{positive} iff it is necessarily possessed by every God-like being.
$$
P_B(\varphi) \equiv \nec \all x. [G_B(x) \imp \varphi(x)]
$$
\end{definition}

Bj{\o}rdal refines the notion of essence (which he calls maximal composite) by restricting the attention to positive properties only. He calls this refined notion of essence a maximal composite.

\begin{definition}[Changed]
\label{B:D2}
A \emph{maximal composite} of an individual's positive properties is a positive property possessed by the individual and necessarily implying every positive property possessed by the individual.
$$MCP(\varphi, x) \equiv (\varphi(x) \wedge P_B (\varphi)) \wedge \all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(\varphi(y) \imp \psi(y)))
$$
\end{definition}

\begin{definition}[Changed]
\label{B:D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its maximal composites.
$$
NE_B (x) \equiv \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))
$$
\end{definition}

\setcounter{axiom}{0}
\begin{axiom}[Changed]
\label{A:A1'} If a property is positive, its negation is not positive: 
$$
\all \varphi. [P_B (\varphi) \imp \neg P_B (\neg\varphi)]
$$
\end{axiom}

\setcounter{axiom}{4}
\begin{axiom}
\label{A:A5'} Necessary existence is a positive property. 
$$
P_B (NE_B )
$$
\end{axiom}

The proof of the possible instantiation of $G_B$ is adjusted (as shown below) by proving Theorem 1 of section \ref{sec:newproof} directly for $G_B$ and not for positive properties in general.

\begin{theorem}
\label{T1}
If God-likeness is a positive property, then it is possibly exemplified:
$$
P(G_B) \imp \pos \ex x.G_B(x)
$$
\end{theorem}
\begin{proof} \hfill

\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.G_B(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.G_B(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: G_B(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.G_B(x) $} \RightLabel{$\imp_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg G_B(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  G_B(x) \imp \neg G_B(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(G_B(x) \imp \neg G_B(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(G_B(x) \imp \neg G_B(x)) $}\RightLabel{$\wedge_I$}
\UIC{$P(\neg G_B)$}
\AXC{Axiom 1} \dashedLine
\UIC{$\all \varphi. [P(\neg G_B) \imp \neg P(G_B)]$} \RightLabel{$\all_E$}
\UIC{$ P( \neg G_B(x)) \imp \neg P(G_B) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(G_B) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(G_B)$} 
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.G_B(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.G_B(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(\rho) \imp \pos \ex x.G_B(x) $} \RightLabel{$\all_I$}
\UIC{$ \all \varphi. [P(G_B) \imp \pos \ex x.G_B(x)] $}

\end{prooftree}
\end{proof}

ToDo: State corollary and prove it, instead of making a forward reference to lemma L1. Then use the corollary in lemma 1.

Corollary 1 of section \ref{sec:newproof} follows by proving the positivity of godlikeness as in lemma \ref{B:L1} below.

\begin{lemma}
\label{B:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G_B(z) \imp \nec \ex x. G_B(x)
$$
\end{lemma}
\begin{proof}
\begin{small}
\begin{prooftree}
\AXC{$ $}
\UIC{$G_B(x) $} 
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. [G_B(x) \imp G_B(x)] $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $}\RightLabel{ $\wedge_I$}
\BIC{$G_B(x) \wedge P_B(G_B) $}
\end{prooftree}
\end{small}

@@

\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A5'}}\dashedLine
\UIC{$P_B (NE_B)$} \RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp NE_B(x)]$}\RightLabel{ D\ref{B:D3}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\nec_E$}
\UIC{$\all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\forall_E$}
\UIC{$ [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}
\AXC{$ $} \RightLabel{ $2$}
\UIC{$G_B(x) $}  \RightLabel{ $\imp_E$}
\BIC{$\all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$} \RightLabel{ $\forall_E$}
\UIC{$(MCP(G_B, x) \imp \nec \ex y.G_B(y))]$}
\AXC{$ $}\dashedLine
\UIC{$MCP(G_B, x) $} \RightLabel{ $\imp_E$}
\BIC{$\nec \ex y.G_B(y)$}\RightLabel{ $\imp_I^2$}
\UIC{$G_B(x) \imp \nec \ex y.G_B(y)$}
\end{prooftree}
\end{small}
\end{proof}

@@

\bibitem[{\itshape Benzm\"uller \& Weber \& Woltzenlogel Paleo (2016)}]{controversy}
Benzm\"uller, C. \& Weber, L. \& Woltzenlogel Paleo, B. 2016. {\itshape Computer-Assisted Analysis of the Anderson-H\'ajek Ontological Controversy}. In: {\itshape Logica Universalis} 10 (2016)

@@

\bibitem[{\itshape Magari(1988)}]{magari}
Magari, R. 1988. {\itshape Logica e Teofilia}. In: {\itshape Notizie di Logica} VII.4.
